---
title: "README"
author: "George G. Vega Yon"
date: "September 9, 2015"
output: 
  md_document:
    variant: markdown_github
---

# netdiffuseR: Network Analysis for Diffusion of Innovations

[![Build Status](https://travis-ci.org/USCCANA/netdiffuseR.svg?branch=master)](https://travis-ci.org/USCCANA/netdiffuseR) [![Build status](https://ci.appveyor.com/api/projects/status/6u48wgl1lqak2jum?svg=true)](https://ci.appveyor.com/project/gvegayon/netdiffuser) [![codecov.io](https://codecov.io/github/USCCANA/netdiffuseR/coverage.svg?branch=master)](https://codecov.io/github/USCCANA/netdiffuseR?branch=master)

This package contains functions useful for analyzing network data for diffusion of innovations applications.

The package was developed as part of the paper Thomas W. Valente, Stephanie R.
Dyal, Kar-Hai Chu, Heather Wipfli, Kayo Fujimoto, *Diffusion of innovations theory applied to global tobacco control treaty ratification*,
Social Science & Medicine, Volume 145, November 2015, Pages 89-97, ISSN 0277-9536
(available [here](http://www.sciencedirect.com/science/article/pii/S027795361530143X))

From the description: Statistical analysis, visualization and simulation of
  network diffusion of innovations. The package implements algorithms for
  calculating stats such as innovation threshold levels, infectiousness
  (contagion) and susceptibility, and hazard rates as presented in Burt (1987),
  Valente (1995), and Myers (2000) (among others).
  
__Acknowledgements__:  netdiffuseR was created with the support of grant R01 CA157577 from the National Cancer Institute/National Institutes of Health. 

## Installation
Infectiousness and Susceptibility


Using the `devtools` package, you can install `netdiffuseR` dev version as follows

```r
devtools::install_github('USCCANA/netdiffuseR', build_vignettes = TRUE)
```

You can skip building vignettes by setting `build_vignettes = FALSE` (so it is not required).

For the case of OSX users, there seems to be a problem when installing packages 
depending on `Rcpp`. This issue, developed [here](https://github.com/USCCANA/netdiffuseR/issues/3),
can be solved by open the terminal and typing the following

```sh
curl -O http://r.research.att.com/libs/gfortran-4.8.2-darwin13.tar.bz2
sudo tar fvxz gfortran-4.8.2-darwin13.tar.bz2 -C /
```

before installing the package through `devtools`.

## Examples

This example has been taken from the package's vignettes:


```{r}
library(netdiffuseR)
```

### Infectiousness and Susceptibility

```{r plot_infectsuscept}
# Generating a random graph
set.seed(1234)
n <- 100
nper <- 20
graph <- rgraph_er(n, nper, .5)
toa <- sample(c(1:(1+nper-1), NA), n, TRUE)
head(toa)

# Creating a diffnet object
diffnet <- as_diffnet(graph, toa)
diffnet
summary(diffnet)

# Visualizing distribution of suscep/infect
out <- plot_infectsuscep(diffnet, bins = 20,K=5, logscale = TRUE)
```

### Threshold

```{r, plot_threshold, fig.height=7}
# Generating a random graph
set.seed(123)
diffnet <- rdiffnet(50, 20, rgraph.args = list(m=4))

# Threshold with fixed vertex size
plot_threshold(diffnet)

# Threshold with vertex size = avg degree
cex <- rowMeans(dgr(diffnet))
cex <- (cex - min(cex) + 1)/(max(cex) - min(cex) + 1)/2
plot_threshold(diffnet, vertex.cex = cex)
```

### Adoption rate

```{r}
plot_adopters(diffnet)
```


### Hazard rate

```{r}
hazard_rate(diffnet)
```


### Diffusion process

```{r plot_diffnet}
plot_diffnet(diffnet, vertex.cex = 2, slices=c(1,7,15,20))
```


## To-do list

- Import/Export functions for interfacing other package's clases, in particular: `statnet` set (specially the packages `networkDynamic` and `ndtv`), `igraph` and `Rsiena`.
- Populate the tests folder.
- Use spells? (`select_egoalter` would use this)
- Classify individuals by adoption category using early adopters, adopters, and laggards, and by threshold using very low, low, high and very high threshold (Valente 95' p. 94).
- Double check all functions using adjacency matrix values.

