---
title: "Epidemiological analysis on diffusion networks"
author: "Anibal Olivera M."
date: "May 14, 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Epidemiological analysis on diffusion networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "85%",
  fig.align = "center"
)
set.seed(2026)
library(netdiffuseR)
```

`netdiffuseR` was built to study how *behaviours* spread through social
networks. Since version 1.26.0 it also speaks the language of
*contagion*: the package ships an epidemiological extension that turns
any diffusion network into a first-class epidemic object, with the
infrastructure needed to compute reproduction numbers, secondary attack
rates, generation times, survival curves, and offspring distributions
directly from the contact graph.

This vignette has two halves. **Part I** is a methods tour: each new
piece of machinery is introduced with a minimal simulated example so
the concept lands before the syntax. **Part II** is a case study on
real experimental data --- the *Epigames* pandemic simulation
(Musa *et al.*, 2026) --- where 594 students played a 15-day infection
game on their campus. The dataset is shipped with the package, so the
analysis is reproducible end-to-end from a fresh install.

The intended reader is a social-network researcher who is comfortable
with `diffnet` objects but new to compartmental / contagion analysis.
The companion vignettes "Introduction to netdiffuseR", "Simulating
diffusion networks" and "Multiple behaviours" cover the substrate this
vignette builds on.

------------------------------------------------------------------------

# Part I --- Methods

## §1 From diffusion to contagion

A classic netdiffuseR question is *"who adopted a new practice, when,
and through which ties?"*. The answer lives in a `diffnet` object,
which records adoption times (`$toa`), the contact network at each
period (`$graph`), and derived state (`$cumadopt`, `$adopt`).

A classic epidemiological question is *"who infected whom, with what
probability, and how fast did the epidemic move?"*. The substrate is
the same --- a graph plus per-node infection times --- but the answer
also requires a *transmission tree*: a who-infected-whom edge list
that links each case to its presumed source. With that tree you can
compute the empirical reproduction number, secondary attack rate, and
generation time --- the standard ingredients of an outbreak report.

netdiffuseR's epidemiological subsystem is exactly the set of
primitives that turn the first object into the second.

## §2 The `diffnet_epi` subclass

A `diffnet_epi` is a `diffnet` with one extra slot: `$transmission`,
holding the tree. The class hierarchy is

``` r
class(x)
#> c("diffnet_epi", "diffnet")
```

so every method that worked on a plain `diffnet` (`plot`, `summary`,
`hazard_rate`, `threshold`, `exposure`, ...) keeps working
unchanged on the subclass via S3 dispatch.

There are three ways to obtain a `diffnet_epi`:

1.  Let `rdiffnet()` build one for you by passing a
    `source_attribution =` callback (it tracks lineage during
    simulation).
2.  Promote an existing `diffnet` from a tree you already have
    (e.g. from another tool or an observed outbreak), via
    `as_transmission_tree()`.
3.  Reconstruct the tree from a plain `diffnet`'s graph and adoption
    times by source-attribution, via
    `as_diffnet_epi(dn, attribution = ...)`.

We start with path (1) because it is the shortest:

```{r quickstart}
set.seed(2026)
dn <- rdiffnet(
  n  = 60, t = 6, seed.graph = "small-world",
  seed.p.adopt = 0.05, stop.no.diff = FALSE,
  threshold.dist = 1, exposure.args = list(normalized = FALSE),
  source_attribution = source_attribution_uniform
)
class(dn)
```

Two of those arguments deserve attention because they recur
throughout this part. `threshold.dist = 1` combined with
`exposure.args = list(normalized = FALSE)` configures `rdiffnet()`
as the simplest possible *virus model*: exposure is the raw count
of infectious neighbours, and the adoption rule is the deterministic
step `exposure >= 1`. Concretely: **a single infectious contact, of
any duration, infects the target with certainty**. The decision is
binary --- if at least one neighbour is infected, the target becomes
infected next period; otherwise it stays susceptible. Edge weights
don't enter the picture under the default identity link and
`valued = FALSE`. This is the deliberately simple null model we
will work with throughout §3-§7. The unrealism of "any duration is
enough" --- real airborne transmission depends non-linearly on
exposure time --- is exactly what §8.1 addresses with the
Wells-Riley equation. In the meantime, every toy in this part uses
the same two arguments.

The result is a `diffnet_epi` (the `source_attribution` argument
is what asks the engine to track infectors as it goes during the
simulation). The transmission tree is now accessible:

```{r tree-shape}
tr <- transmission_tree(dn)
head(tr)
nrow(tr)
```

Each row is one infection event: `target` was infected on `date` by
`source` (which itself entered I on `source_exposure_date`). Seeds
have `source = NA`. The `virus_id` / `virus` columns are useful when
multiple behaviours are tracked at once.

## §3 Source attribution: three bundled rules

In observed data a target can have several already-infected
neighbours; the experiment does not tell you which one transmitted.
Source attribution is the rule used to break that tie. netdiffuseR
ships three kernels (all exported, all follow the same contract so
users can write their own):

| Rule                        | Behaviour                                           |
|-----------------------------|-----------------------------------------------------|
| `source_attribution_uniform`   | Sample one infectious neighbour uniformly at random |
| `source_attribution_weighted`  | Sample with probability proportional to edge weight |
| `source_attribution_earliest`  | Pick the neighbour with the earliest adoption time  |

Pass any of them to `rdiffnet()`, or to the post-hoc reconstruction
function `transmission_tree_from_events()` introduced in §7. The choice affects
*which* node gets credit for each transmission, but not the *total*
number of transmissions: every non-seed adopter contributes one row
to the tree regardless of who is named as their infector. So a
non-obvious algebraic invariant emerges:

```{r attribution-comparison}
make_dn <- function(attr_fn) {
  set.seed(2026)
  rdiffnet(n = 60, t = 6, seed.graph = "small-world",
           seed.p.adopt = 0.05, stop.no.diff = FALSE,
           threshold.dist = 1, exposure.args = list(normalized = FALSE),
           source_attribution = attr_fn)
}

dn_unif <- make_dn(source_attribution_uniform)
dn_earl <- make_dn(source_attribution_earliest)

# Aggregate R is invariant: sum(offspring) = #non-seed edges, fixed by
# (graph, toa) alone, and N (= #cases) is also fixed.
c(uniform_R  = attr(repr_number(dn_unif), "global"),
  earliest_R = attr(repr_number(dn_earl), "global"))
```

What *does* change is the per-source distribution of offspring
counts. Under `_earliest` the earliest infectors collect all the
credit, producing a more concentrated right tail; `_uniform` spreads
it out:

```{r attribution-shape, fig.height=3.5, fig.cap="Same dataset, same aggregate R, different shape of offspring distribution under two attribution rules."}
op <- par(mfrow = c(1, 2))
plot(repr_number(dn_unif),  main = "uniform")
plot(repr_number(dn_earl), main = "earliest")
par(op)
```

Use `_earliest` when you want a deterministic, "first contact wins"
baseline; `_weighted` when contact intensity is meaningful; `_uniform`
when you have no prior. None of them is "the right answer" in an
absolute sense --- they are explicit modelling choices that move
mass between nodes but conserve the global epidemic accounting.

## §4 The five epidemiological metrics

Once a tree is in place, netdiffuseR exposes five tree-aware metrics.
All return classed `data.frame`s with custom `print` methods, so the
console reads as a summary while the full per-row breakdown remains
accessible via standard subscripting / `as.data.frame()`.

```{r quickstart-metrics, eval=TRUE}
pp <- peak_prevalence(dn)
pt <- peak_time(dn)
sar <- secondary_attack_rate(dn)
gt  <- generation_time(dn)
sc <- survival_curve(dn)

cat(sprintf("Peak prevalence    : %.3f at t = %d\n", pp, pt))
sar
gt
sc
```

A quick glossary, since the names overlap with several literatures:

-   **Peak prevalence** is the maximum share of currently-adopted
    nodes across the time horizon. For an absorbing diffusion (no
    recovery, the default) this is the final share; under SIRS it can
    occur mid-run.
-   **Secondary attack rate** (SAR) is `n_secondary / n_contacts`,
    keyed per infection event of the source (one row per
    `(source, virus_id, source_exposure_date)`). The aggregate is
    the sum-over-sum: `sum(n_secondary) / sum(n_contacts)`.
-   **Generation time** is, per tree edge, `date - source_exposure_date`:
    how long the infector had been in state I before transmitting.
-   **Survival curve** is a Kaplan-Meier estimate of "time until
    recovery" among adopters. For absorbing diffusions every adopter
    is right-censored (no recovery ever observed) and the curve is
    flat at 1.

`summary.diffnet_epi()` rolls all five up into a single block on top
of the standard `summary.diffnet` output:

```{r summary-epi}
summary(dn)
```

## §5 Reproduction number and the offspring distribution

In the SIR convention, a node is **susceptible (S)** before it adopts,
**infectious (I)** during the period in which it can transmit, and
**recovered (R)** once it stops. The *reproduction number*
\(R\) is, in plain words, the **average number of secondary cases
that a typical infectious node causes during its time in I**. For an
absorbing diffusion every adopter stays in I until the simulation
ends; under SIRS (see §6) a node can enter and leave I more than
once, and each such infection-life counts as its own case.

The arithmetic is straightforward. Let \(\nu_i\) (the *offspring
count*) be the number of secondary cases attributable to infection
event \(i\) --- exactly the number of times that node-event appears
as a `source` in the transmission tree. The empirical reproduction
number is

\[
R \;=\; \frac{1}{N}\sum_{i=1}^{N} \nu_i,
\]

where \(N\) is the total number of infection events
(seeds + secondary). The convention matches Lloyd-Smith *et al.*
(2005) and `epiworldR::get_reproductive_number()`: seeds are
included as cases (with their offspring counted from the moment
they entered I), and terminal cases (\(\nu_i = 0\)) stay in the
denominator.

```{r repr-toy}
R <- repr_number(dn)
R
```

Three things to notice. First, the printed scalar is the aggregate
mean. Second, the full per-case offspring is one column of the
returned `data.frame` --- ready for plotting, model fitting, or
distributional checks. Third, the `plot` method renders the offspring
distribution as a barplot, the visual signature of superspreading
when the right tail is heavy:

```{r repr-plot, fig.cap="Offspring distribution on a toy absorbing diffusion."}
plot(R)
```

### A note on multi-behaviour datasets

When the tree carries more than one virus, the print method makes the
aggregation explicit and adds a per-diffusion rollup, so the reader
never confuses a pooled `R` with a single-diffusion one:

```{r repr-multi}
set.seed(2026)
suppressMessages(
  dn_multi <- rdiffnet(
    n = 60, t = 6, seed.graph = "small-world",
    seed.p.adopt = list(0.05, 0.03), stop.no.diff = FALSE,
    threshold.dist = 1, exposure.args = list(normalized = FALSE),
    source_attribution = source_attribution_uniform
  )
)
repr_number(dn_multi)
```

The `source_attribution` argument also accepts a *list* of length
\(Q\) (number of diffusions), one attribution rule per virus. This
is useful when the different processes have different plausible
infectors --- e.g. a uniform-attribution behaviour spreading
alongside an earliest-attribution behaviour:

```{r repr-multi-mixed}
set.seed(2026)
suppressMessages(
  dn_mixed <- rdiffnet(
    n = 60, t = 6, seed.graph = "small-world",
    seed.p.adopt = list(0.05, 0.03), stop.no.diff = FALSE,
    threshold.dist = 1, exposure.args = list(normalized = FALSE),
    source_attribution = list(
      source_attribution_uniform,    # for virus 1
      source_attribution_earliest    # for virus 2
    )
  )
)
repr_number(dn_mixed)
```

## §6 SIRS: re-infection support

So far every example was *absorbing*: once a node adopted, it stayed
adopted. To model recovery and re-infection (the S-I-R-S regime), pass
`disadopt =` to `rdiffnet()` --- any function that decides who returns
to S each period. netdiffuseR's tree machinery records each fresh
adoption as its own row, so a node infected three times during the
run shows up as three cases:

```{r sirs-toy}
# disadopt_30 returns the set of nodes that recover each period. Here it
# randomly disadopts 30% of the currently-adopted population at every
# step, mimicking a short infectious window. Any user function with
# signature (expo, cumadopt, time) -> list-of-disadopter-vectors works.
disadopt_30 <- function(expo, cumadopt, time) {
  q_max <- dim(cumadopt)[3]
  res <- vector("list", q_max)
  for (q in seq_len(q_max)) {
    adopters <- which(cumadopt[, time, q] == 1L)
    res[[q]] <- if (length(adopters))
      sample(adopters, ceiling(0.30 * length(adopters))) else integer()
  }
  res
}

set.seed(2026)
dn_sirs <- rdiffnet(
  n = 60, t = 10, seed.graph = "small-world",
  seed.p.adopt = 0.05, stop.no.diff = FALSE,
  threshold.dist = 1, exposure.args = list(normalized = FALSE),
  disadopt = disadopt_30,
  source_attribution = source_attribution_uniform
)
tr_sirs <- transmission_tree(dn_sirs)

# Some nodes appear as targets in more than one row -> re-infected.
node_lives <- table(paste(tr_sirs$target, tr_sirs$virus_id, sep = "::"))
table(as.integer(node_lives))
```

How to read this table: the names are *number of infection events*
that a single `(node, virus_id)` pair experienced over the
simulation; the values are *how many nodes* fall into each bucket.
For an absorbing simulation the table is always `1 -> N` (every
node-virus pair appears in the tree exactly once). Under SIRS,
buckets `2`, `3`, ... light up because some nodes cycle through I
multiple times --- a node in bucket `3` was infected, recovered,
re-infected, recovered again, and re-infected a third time. Each
entry into I is its own *case* for the purposes of `repr_number()`
and `secondary_attack_rate()` (per-event keying), which is the
quantity Lloyd-Smith superspreader analyses operate on.

A subtle consequence worth flagging: under SIRS, a single source can
re-infect the *same* target multiple times during its infection
life. Each transmission is a row in the tree and contributes to
`n_secondary`. The per-row SAR can therefore exceed 1; this is
expected (the unit shifts from "probability per contact" to
"transmissions per contact"), and the aggregate sum-over-sum
interpretation is preserved.

## §7 Reconstructing the tree from observed data

So far every `diffnet_epi` has come straight out of `rdiffnet()` ---
the simulation engine built the tree as it went. In the empirical
setting things look different: you start with a plain `diffnet`
(contact graph + adoption times) and the tree has to be inferred
*post hoc*. The primitive for that is
`transmission_tree_from_events()`:

```{r post-hoc}
# Pretend dn was built by hand from observed data; strip the diffnet_epi
# class so we are working from a plain diffnet.
class(dn) <- "diffnet"

# Reconstruct the tree from x's graph slices + toa, using uniform
# attribution and a fixed seed for reproducibility.
tree <- transmission_tree_from_events(
  dn,
  attribution = "uniform",
  seed        = 2026
)
head(tree)
```

The same job, wrapped in one step that returns the promoted object,
is `as_diffnet_epi(dn, attribution = ...)`:

```{r as-diffnet-epi-reconstruction}
dn_epi <- as_diffnet_epi(dn, attribution = "uniform", seed = 2026)
is.diffnet_epi(dn_epi)
identical(transmission_tree(dn_epi), tree)
```

This is exactly the primitive Part II uses --- the bundled
`epigamesDiffNet` was constructed by running
`as_diffnet_epi(., attribution = "uniform")` on the empirical contact
graph and adoption times. The same algorithm therefore powers both
*"simulate a network outbreak"* and *"analyse a real one"*. Custom
attribution rules are user-functions with the
`source_attribution_uniform` signature, so domain-specific priors
(e.g. proximity duration, role homophily) can be wired in without
touching the package internals.

## §8 Stochasticity: virus exposure and behavioural adoption

Every toy in §3-§7 used the deliberately simple model laid out in
§2: `threshold.dist = 1` with `exposure.args = list(normalized = FALSE)`
--- a node becomes infected the moment one infectious neighbour
exists, with certainty, regardless of edge weight or contact
duration. The package supports **two independent generalisations
of this model**, each one relaxing a different unrealistic
assumption. They sit at different layers of the simulation:

```
   contact graph + adopted neighbours
              |
              v  (a) noisy exposure  ---- exposure.mode = "stochastic"
              |                            exposure.args$link_fun = ...
       exposure value
              |
              v  (b) noisy adoption  ---- adoption_mechanism = adoptmech_logit
              |                            adoption_pars = list(beta0=, beta_expo=)
       adoption decision
```

### §8.1 From threshold to infection probability (Wells-Riley)

So far we have assumed something epidemiologically implausible:
that **a single infectious contact, regardless of its duration,
infects the target with certainty**. A 1-second brush past an
infected person and a 4-hour conversation are treated identically
--- both cross `threshold = 1` because both contribute one unit of
exposure under the default identity link. Real airborne
transmission is neither deterministic nor exposure-duration-blind:
the probability that a given contact transmits *grows with
contact duration but saturates* --- doubling the time you share a
room with an infected person less than doubles your chance of
getting sick. Wells (1955) and Riley, Murphy and Riley (1978)
formalised this in the **Wells-Riley equation**, where the
per-contact probability of infection is

\[
P(\text{infection} \mid \text{contact of duration } w) \;=\; 1 - \exp(-\beta\,w),
\]

with \(\beta\) aggregating the infector's quanta-emission rate,
ventilation, and the target's susceptibility. The function is
linear in \(w\) only at very short durations and saturates at 1 as
\(w \to \infty\) --- exactly the qualitative behaviour we want.
netdiffuseR's `exposure.mode = "stochastic"` with
`link_fun = "wells-riley"` implements this transform on every edge
weight: each non-zero contact becomes a Bernoulli draw with the
saturating probability above. Two changes from the §3-§7 default
work together: contacts are now *probabilistic* (not all transmit)
and edge weights *matter* (longer contacts are riskier).

```{r stoch-exposure}
set.seed(2026)
suppressMessages(
  dn_wr <- rdiffnet(
    n = 60, t = 6, seed.graph = "small-world",
    seed.p.adopt = 0.05, stop.no.diff = FALSE,
    threshold.dist = 1,
    exposure.mode  = "stochastic",
    exposure.args  = list(normalized = FALSE,
                          link_fun  = "wells-riley",
                          link_pars = list(beta = 0.5)),
    source_attribution = source_attribution_weighted
  )
)
cat(sprintf("Final prevalence : %.3f\n", peak_prevalence(dn_wr)))
cat(sprintf("R                : %.3f\n",
            attr(repr_number(dn_wr), "global")))
```

`beta` is the only knob: small `beta` damps transmission per
contact, large `beta` recovers the deterministic 1-contact-infects
regime. The exposure layer is now noisy, but the adoption decision
itself remains deterministic given the (random) exposure count.

### §8.2 From threshold to a noisy behavioural decision (logit on mask adoption)

The virus is not the only thing that diffuses through a contact
network during a pandemic. **Protective behaviours** --- wearing a
mask, accepting a vaccine, voluntarily quarantining --- also spread
person-to-person, and the standard model for them is *adoption*,
not infection. Threshold-based diffusion (§3-§7) is one option;
*probabilistic adoption* is the natural alternative. Each susceptible
individual evaluates their exposure to mask-wearing peers and adopts
with probability

\[
P(\text{wears mask today} \mid \text{exposure}) \;=\; \mathrm{logit}^{-1}\!\bigl(\beta_0 + \beta_{\text{expo}}\,\cdot\,\text{exposure}\bigr).
\]

A negative \(\beta_0\) keeps baseline adoption rare; a positive
\(\beta_{\text{expo}}\) makes adoption increasingly likely as peer
adoption rises. `adoptmech_logit` implements exactly this rule, and
because it overrides `rdiffnet()`'s threshold mechanism, we no
longer need `threshold.dist`:

```{r stoch-adopt}
set.seed(2026)
dn_mask <- rdiffnet(
  n = 60, t = 6, seed.graph = "small-world",
  seed.p.adopt = 0.05, stop.no.diff = FALSE,
  adoption_mechanism = adoptmech_logit,
  adoption_pars      = list(beta0 = -2, beta_expo = 4),
  source_attribution = source_attribution_uniform
)
cat(sprintf("Final mask prevalence : %.3f\n", peak_prevalence(dn_mask)))
cat(sprintf("R (mask adoption)     : %.3f\n",
            attr(repr_number(dn_mask), "global")))
```

The vocabulary deliberately shifts here: we say *adoption*, not
*infection*, because the underlying process is behavioural, not
contagious. The Lloyd-Smith machinery from §5 still applies ---
\(R\) is now the mean number of *secondary mask adopters* per
adopter --- but the interpretation tracks the substantive question.

### §8.3 Together: the simulation analogue of a real pandemic experiment

(a) and (b) are independent and composable. A simulation can have
Wells-Riley exposure for the virus *and* logit adoption for masks
running on the same contact graph as two separate diffusion
processes. That combination --- a stochastically transmitted virus
plus an adoption decision that itself spreads through the same
social network --- is the simulation analogue of a real-world
pandemic experiment. **And that is exactly the setting Epigames
captures in the wild**: a virus propagating through Bluetooth
proximity contacts, while participants make mask / medicine /
isolation decisions whose own contagion the experiment also
records. Part II turns to that data.

------------------------------------------------------------------------

# Part II --- Case study: Epigames

## §9 The experiment

The bundled `epigamesDiffNet` records a controlled pandemic
simulation in which 594 students used a mobile app for 15 days on
their university campus. The app turned face-to-face encounters
(measured by Bluetooth proximity) into a contact graph, and a
seeded *virus* spread through that graph following experimental
parameters (Musa *et al.*, 2026). The dataset shipped here aggregates
the original raw history to daily slices and reconstructs the
transmission tree by uniform source attribution --- which means the
analysis below treats the dataset as a single observed outbreak with
an attribution-model uncertainty we will quantify in §11.

The companion hourly dataset, `epigames`, contains the raw inputs
(static attributes, hourly edgelist, dynamic behavioural attributes)
in case the user wants to build a finer-grained diffnet from scratch.

## §10 Loading and inspecting

```{r epigames-load}
data("epigamesDiffNet", package = "netdiffuseR")
class(epigamesDiffNet)
nnodes(epigamesDiffNet)
length(epigamesDiffNet$graph)   # number of daily slices
```

The header line of the print method shows the standard diffnet
summary plus a transmission tree summary line:

```{r epigames-print}
epigamesDiffNet
```

Because `diffnet_epi` inherits from `diffnet`, the usual descriptive
tooling works out of the box. The cumulative adoption curve, for
instance:

```{r epigames-adopters, fig.cap="Cumulative adoption curve for the Epigames outbreak (n = 594, 15 days)."}
plot_adopters(epigamesDiffNet)
```

## §11 The five metrics

```{r epigames-five}
pp  <- peak_prevalence(epigamesDiffNet)
pt  <- peak_time(epigamesDiffNet)
sar <- secondary_attack_rate(epigamesDiffNet)
gt  <- generation_time(epigamesDiffNet)
sc  <- survival_curve(epigamesDiffNet)

cat(sprintf("Peak prevalence    : %.3f at day %d\n", pp, pt))
cat(sprintf("SAR (aggregate)    : %.3f over %d infection events\n",
            attr(sar, "global"), nrow(sar)))
cat(sprintf("Generation time    : mean = %.2f days, median = %.0f days, n_edges = %d\n",
            mean(gt$gen_time), median(gt$gen_time), nrow(gt)))
```

The peak occurs on the last day of the experiment: the outbreak is
still growing at the horizon. The mean SAR of ~0.18 is consistent
with a moderately transmissible airborne agent under university-life
mixing, with substantial heterogeneity across infectors (we will
visualise this in §12). Mean generation time of ~2.2 days is in the
right ballpark for the parameters used in the experiment.

The survival curve printed above is flat at 1 because the bundled
`epigamesDiffNet$status` array only carries adoption events ---
every adopter is right-censored. This is a *preprocessing artefact*
of the current data pipeline, not a property of the experiment. The
underlying `histories.csv` (in the project's `playground/`) contains
295 `outcome` events, including **42 `RECOVERED`**, 62 `DEAD`, and
191 `ESCAPED` transitions, none of which the bundled diffnet
currently wires in. A future revision of `data-raw/epigamesDiffNet.R`
will read those into a non-monotone `$status` array and make
`survival_curve()` and the SIRS branch of `repr_number()` light up
on this dataset.

## §12 The Epigames reproduction number

```{r epigames-R}
R_emp <- repr_number(epigamesDiffNet)
R_emp
```

The aggregate `R` of \~`r round(attr(repr_number(epigamesDiffNet), "global"), 2)`
is below the classical R > 1 outbreak threshold --- consistent with a
diffusion that grew but never went exponential, and with the high
share of seed-like rows (single contacts on day 1) inflating the
denominator. The offspring distribution, however, is highly skewed:

```{r epigames-R-plot, fig.cap="Empirical offspring distribution from the Epigames transmission tree."}
plot(R_emp)
```

A small number of cases account for many secondary infections. This
is the *superspreading signature* Lloyd-Smith *et al.* (2005)
documented in SARS, MERS, and several other respiratory outbreaks. In
the Epigames data it is visible despite the modest sample size and
the controlled environment.

## §13 Empirical vs simulated: same network, varying mechanism

The empirical `R` is one realisation of one attribution policy on
one tree. To contextualise it, we run a counter-factual:
keep the *same contact graph* used by the experiment and let
`rdiffnet()` simulate an absorbing diffusion with default
threshold-based adoption. Both runs see the same network at each
slice; only the adoption rule differs.

```{r epigames-vs-sim}
set.seed(2026)
n_seeds <- sum(epigamesDiffNet$toa == 1, na.rm = TRUE)

# Simulated counter-factual on the SAME contact graph as the experiment.
# When -seed.graph- is a dynamic graph, -t- is inferred from its length, so
# we omit it here to avoid the redundancy warning.
dn_sim <- rdiffnet(
  seed.graph         = epigamesDiffNet$graph,
  seed.p.adopt       = n_seeds / nnodes(epigamesDiffNet),
  stop.no.diff       = FALSE,
  source_attribution = source_attribution_uniform
)

R_sim <- repr_number(dn_sim)

c(empirical = attr(R_emp, "global"),
  simulated = attr(R_sim, "global"))
```

The aggregate scalars are not directly comparable; the *shape* of the
two trees explains why. The empirical tree is dominated by seeds:
`r sum(is.na(transmission_tree(epigamesDiffNet)$source))` of its
`r nrow(transmission_tree(epigamesDiffNet))` rows are day-1 infections
with no prior adopter to attribute to, all of which enter the
Lloyd-Smith denominator with zero offspring. The simulated cascade,
in contrast, starts from a smaller seeded core and grows
diffusion-like through the network --- so its denominator is
correspondingly smaller and its `R` correspondingly larger.

Side by side:

```{r epigames-vs-sim-plot, fig.height=4.5, fig.cap="Empirical (left) vs simulated (right) offspring distributions on the same contact graph. Both use uniform source attribution; the simulation uses threshold-based adoption with a uniform threshold distribution."}
op <- par(mfrow = c(1, 2))
plot(R_emp, main = "Empirical (Epigames)")
plot(R_sim, main = "Simulated (rdiffnet, threshold)")
par(op)
```

What is comparable across the two panels is the *right tail*. The
simulated threshold model reaches offspring counts of
`r max(R_sim$n_offspring)` per infection event; the empirical
distribution caps at `r max(R_emp$n_offspring)`. The threshold
model on this network produces more concentrated transmission events
than the experiment did, even though the contact graph is identical.
A plausible reading is that the experimental subjects' heterogeneous
behaviour --- mask use, medicine, voluntary isolation --- flattened
the right tail relative to a homogeneous-threshold null model.

The takeaway is methodological: because we held the contact graph
fixed and varied only the adoption rule, any difference in offspring
distributions is attributable to behaviour, not to the network.

------------------------------------------------------------------------

# Appendix

## Related vignettes

-   *Introduction to netdiffuseR* --- the substrate
    (`diffnet`, adoption, exposure).
-   *Simulating diffusion networks: Using the `rdiffnet` function* ---
    detailed coverage of the simulation API, including the
    `adoption_mechanism`, `disadopt`, and exposure modes used here.
-   *Multiple behaviours* --- multi-virus / multi-behaviour diffusion,
    the substrate for the per-diffusion `repr_number` rollup shown in §5 of
    §5.

## References

-   Lloyd-Smith, J. O., Schreiber, S. J., Kopp, P. E., & Getz, W. M.
    (2005). Superspreading and the effect of individual variation on
    disease emergence. *Nature*, 438, 355--359.
    <https://doi.org/10.1038/nature04153>
-   Wells, W. F. (1955). *Airborne Contagion and Air Hygiene: An
    Ecological Study of Droplet Infections.* Harvard University Press.
-   Riley, E. C., Murphy, G., & Riley, R. L. (1978). Airborne spread
    of measles in a suburban elementary school. *American Journal of
    Epidemiology*, 107(5), 421--432.
    <https://doi.org/10.1093/oxfordjournals.aje.a112560>
-   Wallinga, J., & Teunis, P. (2004). Different epidemic curves for
    severe acute respiratory syndrome reveal similar impacts of
    control measures. *American Journal of Epidemiology*, 160(6),
    509--516.
-   White, L. F., & Pagano, M. (2008). A likelihood-based method for
    real-time estimation of the serial interval and reproductive
    number of an epidemic. *Statistics in Medicine*, 27, 2999--3016.
-   Valente, T. W. (1995). *Network Models of the Diffusion of
    Innovations*. Hampton Press.
-   Musa, S.S., Mkandawire, W., Inekwe, T. et al. App-based epidemic 
    game in a university campus reveals how risk perception and 
    behavioral interventions shape disease transmission dynamics. 
    *Sci Rep* (2026). <https://doi.org/10.1038/s41598-026-49530-y>

## Session info

```{r session, echo=FALSE}
sessionInfo()
```
