---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Preliminaries

We are interested in running a large scale simulation featuring hundreds of thousands (if not millions) of vertices. Before we proceed, you need to consider the following:

1. Networks in netdiffuseR are sparse-matrices, thus, as long as the network is sparsely connected, it will fit in a regular computer.
2. At the same time, converting a sparse-matrix into a dense matrix at this scale **is not a good idea**, as you computer may crash. This would be the case, for example, if you want to calculate the geodesic matrix.
3. Following the previous point, using `summary(..., skip.moran = FALSE)`--the default behavior--is not a good idea. When dealing with large graphs, set `skip.moran = TRUE` to avoid memory overflow.

# Case 1: Single simulation

Suppose we want to simulate a diffusion process with the following parameters:

- Network type: Small world with parameters $n=200k, k = 10, p = .1$

```{r setp-1, eval=FALSE}
library(netdiffuseR)
```

We no proceed with the simulation

```{r sim-1, eval=FALSE}
ans_sw <- rdiffnet(n = 200, t = 10)
```



