---
title: "diffusiontest updates"
author: "George G. Vega Yon"
date: "November 5, 2015"
output: 
  ioslides_presentation: 
    fig_caption: yes
    logo: ~/Dropbox/usc/cana/cana_logo.png
---

# Function changes
## Function changes: List
- **adjmatbuild** and **adjByTime** -> edgelist_to_adjmat
    - adjmat_to_edgelist
- **selectionFunctionEgoAlter** -> select_egoalter

- **adoptMat** -> adopt_mat
- **toaMat** -> toa_mat

- ExposureCalc -> exposure_cpp
- cumulativeAdopters -> cumulative_adopt_count_cpp
- hazardrate -> hazard_rate_cpp
- threshold -> threshold_cpp

# Other New features
## Other New features: List
- isolated_cpp
- drop\_isolated\_cpp
- rand\_graph\_cpp
- rand\_dyn\_graph_cpp
- sna::degree -> degree_cpp
- sna::sedist ~> struct_equiv_cpp


# So far...
## So far we have

- All but 2 functions rewritten using Rcpp
- Package building
- All functions with basic documentation (missing details)

## Next steps

- Include datasets
- Package tests
- Build demos
- Write vignettes

# Some examples

## Example 1: Creating a random graph

```{r ex1: Random graph, warning=FALSE, message=FALSE}
library(diffusiontest)
set.seed(123)
graph <- rand_graph()
graph
```
----------------

### Converting to an edgelist

```{r, size="tiny"}
adjmat_to_edgelist(graph)
```

--------------

### Plotting (meantime using sna)

```{r ex1 cont, warning=FALSE, message=FALSE}
library(sna)
sna::gplot(graph)
```

---------------

### Computing degree

```{r ex1 Degree, warning=FALSE, message=FALSE}
# Creating a directed graph
set.seed(123)
graph <- rand_graph(n=4,undirected=FALSE)
graph
```

----------------

### Computing degree (cont.)

```{r ex1 Degree cont}
# Comparing degree measurements
 data.frame(
   In=diffusiontest::degree(graph, 0, undirected = FALSE),
   Out=diffusiontest::degree(graph, 1, undirected = FALSE),
   Degree=diffusiontest::degree(graph, 2, undirected = FALSE)
 )
```

---------------------

### Finding and droping isolated vertices

```{r ex1 Isolated}
graph
isolated(graph, undirected=FALSE)

```


## Example 2: Creating a dynamic random graph

<div class="columns-2">
```{r ex2: Dyn Random graph, warning=FALSE, message=FALSE}
set.seed(1234)
graph <- rand_graph(n=5,t=3)
graph

# Vector of adoption times
times <- sample(2001:2003, 5, TRUE)
```
</div>

----------

### Computing exposure: First the adopt mat

```{r ex2 Adopt mat}
adopt <- adopt_mat(times)
adopt
```

---------

### We can see how it looks like

```{r, echo=TRUE, warning=FALSE, message=FALSE}
plot_diffnet(graph, adopt$cumadopt)
```

<p style="font-size:12px;">Adopters: Blue<br>Non Adopters: Grey</p>

-------------

### Computing exposure: Now, the exposure

```{r ex2 Exposure}
exposure(graph, adopt$cumadopt)
```

Note that the function works specifically for arrays

```{r ex2 Exposure err}
e <- tryCatch(exposure(graph[,,1], adopt$cumadopt), 
         error=function(e) e)
```

```{r, echo=FALSE}
structure(e, class="knit_asis")
```

