---
title: |
  | Understanding Diffusion with __netdiffuseR__
  | _Survival Analysis_
subtitle: Sunbelt 2016 INSNA
author: George Vega Yon \and Thomas Valente
date: |
  | Newport Beach, CA
  | April 5, 2016
institute: |
  | Department of Preventive Medicine
  | University of Southern California
header-includes: \usepackage{grffile}
output: beamer_presentation
fontsize: 10pt
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(SparseM)
```

# Setup

We will use the medical innovations data

\scriptsize
```{r setup, include=TRUE}
# Loading the required packages
library(survival)
library(netdiffuseR)

# Loading the data
data("medInnovationsDiffNet")
medInnovationsDiffNet
```
\normalsize

# Preparing the data

From __netdiffuseR__ we will get the following covariates:

- Cohesive exposure: Proportion of ego's adopters at each time period
- Structural equivalence exposure: Same as before but using the structural equivalence graph instead of the baseline network.

\scriptsize
```{r}
# Structural equivalence exposure
medInnovationsDiffNet[["seexp"]] <- exposure(
  medInnovationsDiffNet, groupvar="city", alt.graph = "se")

# Cohesive exposure
medInnovationsDiffNet[["cohexp"]] <- exposure(medInnovationsDiffNet)
```
\normalsize

# Coercing data into a dataframe

\tiny
```{r}
# Getting the data for running the cox regression
dat <- diffnet.attrs(medInnovationsDiffNet, as.df = TRUE)
dat <- subset(
  dat,
  subset = 
    per <= toa & 
    per < 18 &
    !is.na(date),
  select = c(id, per, toa, date, city, proage, proage2, seexp, cohexp))

# Creating the event variable
dat$event <- with(dat, toa==per)

# Here, since its survival, we only care from when the doctor is aware.
dat <- subset(dat, per - date >=0)

# Checking out the data
dat <- dat[with(dat, order(id, per)),]
head(dat,10)
```
\normalsize

Notice that `diffnet.attrs` generates two extra variables: `per` (time period) and `id`.

# The `survival` package

-   In order to work, the `survival` package works with `Surv` objects.
-   These store the response/events and the time frame during which these occurred.
-   Usually take the following form: `Surv(start, end, event)`.
-   For this tutorial we will use the Cox model, from Andersen and Gill (1982)

    $$\mathcal{L}(\beta) = \prod_{i=1}^n\left\{\frac{\exp\beta'x_i(T_i)}{\sum_{j}\exp{\beta'x_j(T_j)}}\right\}^{\delta_i}$$

    Which can extended to time-variant covariates.

# The `survival` package
## The `Suvreg` object

-   First, we need to create the `Suvreg` object using the function with the same name

    \scriptsize
    ```{r}
    # Needs a start, stop, event
    surv_mi <- with(dat, Surv(per-1, per, event))
    ```
    \normalsize
    Notice the warning as the time frames should be grater than 1. 
-   Now, let's take a look at the object itself
    \scriptsize
    ```{r}
    head(surv_mi, 10)
    ```
    \normalsize

# Fitting the model
## All cities

\scriptsize
```{r}
# Fitting a model
set.seed(1988)
mymodel <- formula(surv_mi ~ factor(city) + proage + I(proage^2) +
                     seexp + cohexp + cluster(id))
out <- coxph(mymodel, data=dat)
```

\normalsize

# Fitting the model
## All cities (cont. 1)

\scriptsize
```{r, echo=FALSE}
pander::pander(out)
```
\normalsize

# Fitting the model
## All cities (cont. 2)

More diganostics can be done as follows:

\tiny
```{r out.width=".6\\linewidth"}
# Diagnostics
fit <- survfit(out)

plot(fit, mark.time=TRUE, lty=1:2,
     xlab="Time period", ylab="Survival")
legend("bottomleft", c("Adopters", "Non-adopters"),
       lty=2:1, bty="n")
```
\normalsize


# Example with brazilian farmers

\scriptsize
```{r, out.width='.8\\linewidth'}
data("brfarmersDiffNet")

# Creating a classification for village
village  <- as.integer(factor(brfarmersDiffNet[["village"]])) + 2
nvillage <- length(unique(village))
plot_threshold(brfarmersDiffNet, 
               vertex.sides = village, # Creates polygons
               vertex.col   = terrain.colors(nvillage)[village]) # Colors!
```
\normalsize

# Example with brazilian farmers
## Preparing the data

\tiny
```{r}
# Exposure variables
brfarmersDiffNet[["seexp"]] <- exposure(brfarmersDiffNet, alt.graph = "se",
                                        groupvar="village", valued = TRUE)
brfarmersDiffNet[["cohexp"]] <- exposure(brfarmersDiffNet)

# Creating a dynamic version of age
age <- brfarmersDiffNet[["age"]]
pers <- brfarmersDiffNet$meta$pers
brfarmersDiffNet[["age_dyn"]] <- lapply(
  seq_len(nslices(brfarmersDiffNet)), function(x) {
    age + (pers[x] - 1966) # Surveyed in 1966
  })

# Subset
dat <- diffnet.attrs(brfarmersDiffNet, as.df = TRUE)
dat <- subset(dat, per <= toa, select=c(per, toa, age_dyn, village, seexp, cohexp, id))

# Creating the event variable
dat$event <- with(dat, toa==per)

# Checking out the data
dat <- dat[with(dat, order(id, per)),]
head(dat,10)
```
\normalsize

# Example with brazilian farmers
## Fitting the data

\tiny
```{r}
out <- coxph(Surv(per-1, per, event) ~ factor(village) + seexp + cohexp + age_dyn + cluster(id),
             data=dat)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander::pander(out)
```
\normalsize

# Example with brazilian farmers
## Diagnotstics

More diganostics can be done as follows:

\tiny
```{r out.width=".6\\linewidth"}
# Diagnostics
fit <- survfit(out)

plot(fit, mark.time=TRUE, lty=1:2,
     xlab="Time period", ylab="Survival", firstx = min(dat$per))
legend("bottomleft", c("Adopters", "Non-adopters"),
       lty=2:1, bty="n")
```
\normalsize


# Example with brazilian farmers
## Diagnotstics (cont.)

\tiny
```{r out.width=".6\\linewidth"}
cox.zph(out)
```

\normalsize

