---
title: "Introduction"
author: "Thomas W. Valente and George G. Vega Yon"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")

```

# Diffusion Network Object (diffnet)

- __netdiffuseR__ has its own class of objects: `diffnet`.

- Most of the package's functions accept different types of graphs:
    * Static: `matrix`, `dgCMatrix` (from the __Matrix__ pkg), 
    * Dynamic: `list` + `dgCMatrix`, `array`, `diffnet`
  
- But `diffnet` is the class from which you get the most.

- From __netdiffuseR__'s perspective, network data comes in three classes:

    1. Raw R network data: Datasets with edgelist, attributes, survey data, etc.
    2. Already R data: already read into R using igraph, statnet, etc. (`igraph_to_diffnet`, `network_to_diffnet`, etc.)
    3. Graph files: DL, UCINET, pajek, etc. (`read_pajek`, `read_dl`, `read_ucinet`, etc.)

- In this presentation we will show focus on 1.

# What is a `diffnet` object

A diffusion network, a.k.a. `diffnet` object, is a `list` that holds the following objects:

- `graph`: A `list` with $t$ `dgCMatrix` matrices of size $n\times n$,
- `toa`: An integer vector of length $n$,
- `adopt`: A matrix of size $n\times t$,
- `cumadopt`: A matrix of size $n\times t$,
- `vertex.static.attrs`: A `data.frame` of size $n\times k$,
- `vertex.dyn.attrs`: A list with $t$ dataframes of size $n\times k$,
- `graph.attrs`: Currently ignored..., and
- `meta`: A list with metadata about the object.

These are created using `as_diffnet` (or its wrappers).


# What does netdiffuseR includes

-   Among __netdiffuseR__ features, we find three classical Diffusion Network Datasets:
    
    - `brfarmersDiffNet` Brazilian farmers and the innovation of Hybrid Corn Seed (1966).
    - `medInnovationsDiffNet` Doctors and the innovation of Tetracycline (1955).
    - `kfamilyDiffNet` Korean women and Family Planning methods (1973).
    
    ```{r printing}
    brfarmersDiffNet
    medInnovationsDiffNet
    kfamilyDiffNet
    ```

-   Functions to compute:
    
    *   (complex) Exposure
    *   
