---
title: "Reading Data"
author: "Thomas W. Valente and George G. Vega Yon"
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(comment = "#")
library(netdiffuseR)
library(igraph)
library(networkDynamic)
```

# Basic Diffusion Network

*   To create `diffnet` objects we only need a network and times of adoption:

    ```{r 2-static-net}
    set.seed(9)
    
    # Network
    net <- rgraph_ws(500, 4, .2)
    net <- replicate(5, net, "list")
    
    # Times of adoption
    toa <- sample(c(NA, 2010:2014), 500, TRUE)
    
    diffnet_static <- as_diffnet(net, toa)
    summary(diffnet_static)
    ```

# Static survey

*   netdiffuseR can also read survey (nomination) data:
    
    ```{r 2-static-survey}
    data("fakesurvey")
    fakesurvey
    ```

*   In group one 4 nominates id 6, who does not show in the data, and in group two 1 nominates 3, 4, and 8, also individuals who don't show up in the survey.
    
    ```{r 2-static-survey-cont1}
    d1 <- survey_to_diffnet(
      dat      = fakesurvey,                # Dataset
      idvar    = "id",                      # The name of the idvar
      netvars  = c("net1", "net2", "net3"), # Name of the nomination variables
      groupvar = "group",                   # Group variable (if any)
      toavar   = "toa"                      # Name of the time of adoption variable
      ); d1
    ```
    
*   If you want to include those, you can use the option `no.unsurveyed`
    
    ```{r}
    d2 <- survey_to_diffnet(
      dat      = fakesurvey,
      idvar    = "id",
      netvars  = c("net1", "net2", "net3"),
      groupvar = "group",
      toavar   = "toa",
      no.unsurveyed = FALSE
      ); d2
    ```

*   We can also check the difference
    
    
    ```{r}
    d2 - d1
    rownames(d2 - d1)
    ```

# Dynamic survey

```{r 2-static-dynsurvey}
data("fakesurveyDyn")
fakesurveyDyn
```

```{r 2-static-dynsurvey-cont1}
diffnet_dynsurvey <- survey_to_diffnet(
  dat      = fakesurveyDyn,
  idvar    = "id",
  netvars  = c("net1", "net2", "net3"),
  groupvar = "group",
  toavar   = "toa",
  timevar  = "time"
  )

plot_diffnet(diffnet_dynsurvey)
```

# Other formats

*   The package also supports working with other network formats.

*   Besides of `.net` (Pajek), and `ml` (UCINET), netdiffuseR can actually
    convert between classes: `igraph`, `network`, and `networkDynamic`.
    
        
    ```{r foreign, warnings=FALSE, messages=FALSE}
    dn_ig  <- diffnet_to_igraph(medInnovationsDiffNet) #; dn_ig[[1]]
    dn_net <- diffnet_to_network(medInnovationsDiffNet)
    dn_ndy <- diffnet_to_networkDynamic(medInnovationsDiffNet); dn_ndy
    ```
    
    First two examples it creates a list of objects, the later actually
    creates a single object
      
    ```{r networkDyn}
    networkDynamic_to_diffnet(dn_ndy, toavar = "toa")
    ```

# Problems

2.  Using the rda file _problem2.rda_, read in the edgelist `net_edgelist` and the adjacency
    matrix `net_list` as a diffnet object together with the data.frame `X`.
    
```{r 2-generates-data, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE, results='hide'}
library(igraph)
library(networkDynamic)
# source("diffnet_to_network.r")
set.seed(81231)

# Generating random data
net_list     <- rgraph_er(20, 5)
net_edgelist <- adjmat_to_edgelist(net_list)

X <- data.frame(
  idnum     = rep(1:20, 5),
  var1      = rnorm(20*5),
  YearAdopt = rep(sample(c(NA, 1:5), 20, TRUE), 5),
  year      = sort(rep(1:5, 20))
  )

# Diffnet object
diffnet_0 <- edgelist_to_diffnet(
  edgelist = net_edgelist[,1:2],
  t0       = net_edgelist[, "time"],
  dat      = X,
  idvar    = "idnum",
  toavar   = "YearAdopt",
  timevar  = "year"
)

# Igraph object
igraph_0  <- diffnet_to_igraph(diffnet_0)
network_0 <- diffnet_to_networkDynamic(diffnet_0)


# Storing the data
save(diffnet_0, igraph_0, network_0, X, net_edgelist, net, file = "problem2.rda")
```

3.  With the new diffnet object, apply the same analysis as before.
    Which strategy maximizes adoption?
