---
title: "netdiffuseR showcase: Medical Innovations"
author: "George G. Vega Yon"
date: "January 13, 2016"
output:
  pdf_document:
    keep_tex: true
---

```{r Setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
# opts_chunk$set(out.width = ".7\\linewidth")
```

# Data preprocessing

Reading the data from an stata file.

```{r Loading pkgs and reading the data}
library(foreign)
library(netdiffuseR)

# medInnovations <- read.dta("../../../data-raw/mi_v2.dta")
medInnovations <- read.dta("data-raw/mi_v2.dta")
```

Now that we have the data in R, we can start working with it, in particular, we want to do the following things:

- Create a unique id for each individual in the network.
- Remove unsurveyed individuals (we don't have additional covariates for them).
- Reshaping the data to long format (so we can use it as a panel), and furthermore, as an edgelist.

```{r Preparing data for netdiffuseR}
# Creating unique ids (including for the network data)
othervars <- c("id", "toa", "city")
netvars <- names(medInnovations)[grepl("^net", names(medInnovations))]
for (i in c("id", netvars))
  medInnovations[[i]] <- medInnovations[[i]] + medInnovations$city*1000

# Leaving unsurveyed individuals with NA
surveyed <- medInnovations$id
for (i in netvars)
  medInnovations[[i]][which(!(medInnovations[[i]] %in% surveyed))] <- NA

# Adding autoedges to farmers that are isolated, we need to do this otherwize
# these will be dropped when calling the function -edgelist_to_adjmat-. Notice
# that this does not imply that the graph will have autoedges. (see manual)
isolated <- which(apply(medInnovations[, netvars], 1, function(x) all(is.na(x))))
medInnovations[isolated, netvars[1]] <- medInnovations$id[isolated]

# Reshaping data (so we have an edgelist)
medInnovations.long <- reshape(
  medInnovations[,c(othervars, netvars)], v.names= "net",
  varying = netvars,
  timevar = "level", idvar="id", direction="long")
```


Once we have the data in long format, we can coerce it into an `diffnet` object. This is done by reading the edgelist, obtaining the times of adoption vector and applying the `as_diffnet` function.

```{r Importing data to netdiffuseR}
# Coersing the edgelist to an adjacency matrix. Here we are assuming that the
# network is constant through time.
graph <- with(
  medInnovations.long,
  edgelist_to_adjmat(cbind(id, net), t=18,undirected=FALSE, use.incomplete=FALSE)
)

# Here we are retrieving the set of individuals who actually were used in the
# network (as these are not isolated nodes)
used.vertex <- rownames(graph[[1]])
medInnovations <- subset(medInnovations, id %in% used.vertex)

# Create the vector (subset) of times of adoption using only the individuals
# that are included in the adjacency matrix
toa <- medInnovations$toa

# Creating a diffnet object
diffnet <- as_diffnet(graph, toa)
```

Once a `diffnet` object, we can apply the usual generic R functions:

```{r Checking-the-methods}
plot(diffnet, t=diffnet$meta$nper)
diffnet
summary(diffnet)
```

And the ones included in the package:

```{r graphs, out.width=".7\\linewidth"}
plot_diffnet(diffnet, slices=c(1,4,8,12,16,18))
plot_infectsuscep(diffnet, K=5, logscale = TRUE, bins=20)
plot_threshold(diffnet, undirected = FALSE, vertex.cex = 1/5)
plot_adopters(diffnet)
plot_hazard(diffnet)
```

# Fitting probability of adoption

In this section we use the diffnet object to create a set of variables that we'll be using in a probit model traying to understand the adoption process. First we setup the data for the `glm` function of the package `stats`

```{r}
# Creating a data set
d <- data.frame(
  id = diffnet$meta$ids,
  threshold(diffnet),infection(diffnet, K=3),susceptibility(diffnet, K=3),
  toa=diffnet$toa, indeg=rowMeans(dgr(diffnet, "degree")),
  expo=exposure(diffnet, w=1),
  adopt=toa_mat(diffnet)$cumadopt
)

# Creating dummies per city, home hospital and year
cities <- unique(medInnovations$city)
for (i in sort(cities)) {
  d[[sprintf("city%02d",i)]] <- as.integer(medInnovations$city == i)
}

home <- unique(medInnovations$home)
for (i in sort(home)) {
  d[[sprintf("home%02d",i)]] <- as.integer(medInnovations$home == i)
}

spe <- unique(medInnovations$special)
for (i in sort(spe)) {
  d[[sprintf("spe%02d",i)]] <- as.integer(medInnovations$special == i)
}

month <- unique(medInnovations$toa)
for (i in sort(month)) {
  d[[sprintf("month%02d",i)]] <- as.integer(medInnovations$toa == i)
}

# Joining with other variables
d <- merge(
  d,
  subset(medInnovations, select=c(
    nojourn, proage, friends, id, proage2, social))
  )

# head(d)

# Expanding the data
d.long <- reshape(
  d,
  varying = list(paste("adopt", 1:18,sep="."), paste("expo", 1:18, sep=".")),
  idvar="id", direction="long")

d.long <- d.long[with(d.long, order(id, time)),]
d.long <- subset(d.long, toa >= time)

# head(d.long, 36)
```

\pagebreak

## General model: All cities together (no interaction term)

```{r Adoption Probit}
# Model with all
mod_all_all <- as.formula(paste(
  "adopt.1 ~ expo.1 + nojourn + susceptibility", # proage + proage2 + friends + social
  paste0("city0", 2:4, collapse=" + "),
  paste0(sprintf("month%02d", sort(month)[-1]), collapse=" + "),
  paste0(sprintf("spe%02d", sort(spe)[-1]), collapse=" + "),
  paste0(sprintf("home%02d", sort(home)[-1]), collapse=" + "), 
  sep = " + "
))

out_all_all <- glm(mod_all_all, data=d.long, family = binomial(link="logit"))

# Model without year
mod_all_noyear <- as.formula(paste(
  "adopt.1 ~ expo.1 + nojourn + susceptibility", # proage + proage2 + friends + social
  paste0("city0", 2:4, collapse=" + "),
  # paste0(sprintf("month%02d", sort(month)[-1]), collapse=" + "),
  paste0(sprintf("spe%02d", sort(spe)[-1]), collapse=" + "),
  paste0(sprintf("home%02d", sort(home)[-1]), collapse=" + "), 
  sep = " + "
))

out_all_noyear <- glm(mod_all_noyear, data=d.long, family = binomial(link="logit"))

# Model without year or city
mod_all_noyearnocity <- as.formula(paste(
  "adopt.1 ~ expo.1 + nojourn + susceptibility", # proage + proage2 + friends + social
  # paste0("city0", 2:4, collapse=" + "),
  # paste0(sprintf("month%02d", sort(month)[-1]), collapse=" + "),
  paste0(sprintf("spe%02d", sort(spe)[-1]), collapse=" + "),
  paste0(sprintf("home%02d", sort(home)[-1]), collapse=" + "), 
  sep = " + "
))

out_all_noyearnocity <- glm(mod_all_noyearnocity, data=d.long, family = binomial(link="logit"))
```

```{r Printing output, results='asis'}
texreg::texreg(
  list(out_all_all, out_all_noyear, out_all_noyearnocity), single.row=TRUE,
  custom.model.names = c("Full", "Without year", "Without year and city")
)
```


```{r Threshold OLS, include=FALSE}
## Model 2: Variables affecting threshold
mod2 <- glm(threshold ~
              indeg + city02 + city03 + city04 + proage + nojourn +
              friends + proage2 + social +
#               home02 + home03 + home04 + home06 + home07 + home08 +
#                 home09 + home10 + home11 + home12 + 
              spe02 + spe03 + spe04,
            data=d, family = gaussian())
mod2
```

```{r Susceptibility OLS, include=FALSE}
## Models 3 and 4: Variables affecting susceptibility and ifectiousness
mod3.1 <- glm(susceptibility ~ 
              indeg + expo.1 + city02 + city03 + city04 + proage + nojourn +
              friends + proage2 + social +
#               home02 + home03 + home04 + home06 + home07 + home08 +
#                 home09 + home10 + home11 + home12 + 
              spe02 + spe03 + spe04,
    data=d, family = gaussian())
mod3.1
```

```{r Infection OLS, include=FALSE}
mod3.2 <- glm(infection ~ 
              indeg + expo.1 + city02 + city03 + city04 + proage + nojourn +
              friends + proage2 + social +
#               home02 + home03 + home04 + home06 + home07 + home08 +
#                 home09 + home10 + home11 + home12 + 
              spe02 + spe03 + spe04,
    data=d, family = gaussian())
mod3.1
```

```{r results='asis', include=FALSE}
# texreg::texreg(list(mod1, mod2, mod3.1, mod3.2), 
               # custom.model.names = c("Adoption", "Threshold", "Susceptibility", "Infectiousness") )
```


```{r ergm example, include=FALSE}
# Working with the `ergm` package
library(ergm)

net <- network(as.matrix(diffnet$graph[[18]]))
set.vertex.attribute(net, "threshold", threshold(diffnet)[,1])
set.vertex.attribute(net, "infection", infection(diffnet, K=3)[,1])
set.vertex.attribute(net, "susceptibility", susceptibility(diffnet, K=3)[,1])
set.vertex.attribute(net, "toa", diffnet$toa)


```

```{r RSiena example, include=FALSE}
# Working with the `RSiena` package
library(RSiena)



```


```{r}
library(boot)

# Computing the boot hazard rate
x <- boot(diffnet$cumadopt, function(x,i) hazard_rate(x[i,], no.plot = TRUE), R=1000)
y <- colMeans(x$t)
z <- t(apply(x$t, 2, function(x) quantile(x, c(.05,.5,.95))))
```

